---
title: '{farrago}: an R Package of odds-and-ends'
author: al-obrien
date: '2022-06-01'
slug: farrago
categories: []
tags:
  - R
  - Package
subtitle: ''
summary: ''
authors: [al-obrien]
lastmod: '2022-06-01T12:22:24-07:00'
featured: no
image:
  caption: '[Photo by Todd Quackenbush on Unsplash](https://unsplash.com/photos/IClZBVw5W5A)'
  focal_point: ''
  preview_only: yes
projects: []
codefolding_show: show
---

```{r, include=FALSE}
library(farrago)
library(magrittr)
library(data.table)
library(tibble)

convert_date2fluseason <- function(date, format = '%Y-%m-%d', flu_wk_start = 35, return_values, split_wk53 = TRUE, sunday_start = TRUE){

  valid_returns <- c('week', 'month', 'year', 'season')
  if(!missing(return_values)) match.arg(return_values, valid_returns, several.ok = TRUE)

  # Run through parsing check for YYYY-MM-DD formats
  date <- lubridate::as_date(date, format = format)

  # Parse year, month, wk number, wk day (vectors)
  year_test <- lubridate::year(date)
  month <- lubridate::month(date)
  if(sunday_start) {fmt_wk <- lubridate::isoweek(date + 1)} else {fmt_wk <- lubridate::isoweek(date)} # +1 for Sunday start...
  week_day <- lubridate::wday(date)

  if(split_wk53){
    # Calculate conditions for weeks and season
    fmt_wk[fmt_wk == 53 & month == 12] <- 52 # Roll back to week 52 if in prior year, Dec
    fmt_wk[fmt_wk == 53 & month == 1] <- 1 # Roll into week 1 if in next year, Jan

    # week <- fmt_wk
    # index1 <- week_day == 1 & fmt_wk + 1 == 53 & month == 12 # if Sunday, in Dec & just shy of week 53 (will force to 52)
    # index2 <- !index1 & week_day == 1 # Sunday but not matched to prior index
    # week[index1] <- 52
    # week[index2] <- fmt_wk[index2] + 1
  }
  week <- fmt_wk


  year <- year_test
  index3 <- week==1 & month == 12
  year[index3] <- year_test[index3]+1

  season <- rep(NA, length(fmt_wk))
  index4 <- fmt_wk < flu_wk_start | (fmt_wk == 52 & month == 1) # Go back a year for season for those in 52 wk but next year month
  index5 <- !index4 & fmt_wk == 53 & month == 12
  season[index4] <- paste(year[index4]-1, year[index4], sep="-")
  season[index5 & is.na(season)] <- paste(year[index5 & is.na(season)]+1, year[index5 & is.na(season)]+2, sep="-")
  season[!index5 & is.na(season)] <-  paste(year[!index5 & is.na(season)],year[!index5 & is.na(season)]+1, sep="-")

  out <- setNames(list(week, month, year, season),
                  valid_returns)

  if(!missing(return_values)) {
    return(out[return_values])
  } else {
    return(out)
  }
}
```


After working with R both professionally and through hobby projects I have accumulated an assortment of coding 'odds-and-ends', snippets that I have found useful enough to collate and document. [{farrago}](https://github.com/al-obrien/farrago) is an R package serving as a personal collection of tools to assist with data workflows and analysis, with a focus on health surveillance and epidemiological data; however, it may have utility to other audiences as well.

The functions are organized by general operation type, such as *calculations*, *conversions*, *content creation*, *data transfer*, and *plotting*. As the package continues to grow, I anticipate that particular function themes will be unified under their own distinct packages. This will ensure that the number of dependencies are minimized, making it simpler to maintain. 

Functions that have been topics of [previous posts](/post/data-stowaway/) (e.g. `stow()` and `retrieve()`) are included in {farrago}. Further development and fixes are routinely incorporated in the package (and are not typically deserving of additional blog spotlight). The package content continues to expand and other functions in {farrago} may warrant additional attention in future blog posts. A few of particular interest I have given a cursory introduction below... 

## File transfer

Uploading and downloading files via SFTP is a common operation and, in most instances, it is possible to automate these steps. When working on Windows OS, WinSCP is a common tool used for this purpose and the option of using batch files helps streamline the process. It is a small leap to then have R write and pass these WinSCP batch files to a pre-established configuration on WinSCP. {farrago} can upload and download from these connections after the initial setup in WinSCP through the `transfer_winscp()` function:

```{r, eval = FALSE, echo = TRUE}
drop_location <- 'C:/PATH/TO/DESIRED/FOLDER/'
  
transfer_winscp(file ='my_rmt_file.csv'),
               direction = 'download',
               connection = 'sftp://myusername:mypwd@hostlocation.ca/'
               rmt_path = './location/',
               drop_location = drop_location)

my_transferred_file <- read.csv(paste0(drop_location, 'my_rmt_file.csv'))
```

## Pesky dates

When working with data it is hard to escape working with dates. In several instances they can be frustrating to deal with. 

One of which is determining flu weeks each year... 
```{r}
# Sample date list for flu seasons
date_list <- c('2022-01-01',
               '2021-01-01',
               '2020-08-30',
               '2020-09-01',
               '2020-09-23',
               '2020-01-01',
               '2019-12-31',
               '2018-01-01', 
               '2017-01-01', 
               '2016-01-01')

# Return just the week and season category
convert_date2fluseason(date_list, return_values = c('week', 'season'))
```

Another example is determining holidays, which can be useful to schedule reports...

```{r}
# Find labour day (first monday in Sept) 
list_weekdays(2021, 9, 'Monday')[1]
```

```{r}
# Find observed holiday of xmas day
find_observedday(2021, month = 12, day = 25)
```

## Cohorts

Several operations are useful for working with cohorts but can be difficult to code efficiently in R, especially if they require `for` loops. This includes identifying overlapping time-frames, collapsing cohort entry/exit by a threshold, and assigning episode periods across repeat observations.
```{r, message = FALSE}
test_data <- tibble::tribble(~grp_id, ~date_start, ~date_end,
                      1, '2020-01-01', '2020-01-02',
                      1, '2020-01-03', '2020-01-04',
                      1, '2020-01-04', '2020-09-02',
                      2, '2020-01-01', '2020-09-02',
                      2, '2020-09-10', '2020-09-20',
                      2, '2020-09-21', '2020-09-22',
                      3, '2020-01-01', '2020-01-02',
                      3, '2020-01-02', '2020-01-20',
                      3, '2020-01-21', '2020-01-22',
                      3, '2020-01-22', '2020-04-02',
                      3, '2020-04-22', '2021-04-22',
                      3, '2021-06-09', '2021-06-22') %>%
   dplyr::mutate_at(dplyr::vars(contains('date')), lubridate::ymd)

# Assign an incremental id to rows overlapping (i.e. date end is same or prior to the next start date; these will have same value)
test_data$overlap_id <- identify_overlap(data = test_data,
                                      grp_id = grp_id,
                                      date_start = date_start,
                                      date_end = date_end)

# Create group when difference between date end and start start is 1 day
test_data$collapsed_grp <- collapse_timesteps(data = test_data,
                                              grp_id = grp_id,
                                              date_start = date_start,
                                              date_end = date_end,
                                              threshold = 1)

# Create episode period based upon start dates and a threshold of 10
test_data$episode_period <- assign_episode(data = test_data,
                             grp_id = grp_id,
                             date = date_start,
                             threshold = 10)

knitr::kable(test_data) %>% kableExtra::scroll_box(width = "100%", box_css = 'padding: 5px;')
```

## Create and summarise groups

Although this can be achieved easily with packages like {dplyr}, the `create_hypercube` function makes it simple to create summarised tables that include rows for totals by provided grouping columns.
```{r, message = FALSE}
create_hypercube(mtcars, columns = c('gear', 'cyl'), drop_sum_columns = 'cyl')
```

When working with ages or binning numeric values into discrete categories it is convenient to have automatic labeling...just assign the breaks and let `create_breaks` do the rest.
```{r, message = FALSE}
data = data.frame(val = c(-1,0,10,5,999,9))

breaks = c(0, 1, 10,50,100)

data$cat <- create_breaks(data$val, breaks, TRUE)

data
```

If you have used SAS, you may be familiar with `MULTILABEL FORMAT` summaries. In {farrago}, `create_multilevel_factor` can simulate this kind of behavior and create a dataset with combined category levels. Below shows an example of how overlapping age groups can be added and summarized.
```{r, message = FALSE}
# Sample data with groups and conditions
example_data <- tibble(group = c('Area1', 'Area1', 'Area1', 'Area2', 'Area3', 'Area3'),
                       condition = factor(c('<5', '5-10', '11-20', '<5', '5-10', '20+'), ordered = FALSE))

# Create dataset with additional levels defined
new_data <- create_multilevel_factor(example_data,
                         target_col = 'condition',
                         new_levels = list('NewGrp:0-10' = c('<5', '5-10'), 'NewGrp:5-20' = c('11-20', '5-10')), # Define combined categories
                         collapse = FALSE,
                         track = TRUE)

# Summarise the multi-level grouping by their condition
tmptbl <- addmargins(table(new_data$group,new_data$condition), margin = 1)
knitr::kable(tmptbl)
```

## Wrap-up

If the examples above have kept your attention, I encourage interested readers to explore the package and associated documentation more thoroughly on the repository. Due to the scope of the package, it is simply hosted on [GitHub](https://github.com/al-obrien/farrago). 
